Verification of the CVM algorithm with a Functional Probabilistic Invariant

### 2025-07-04 paper/main.tex
minor fixes to intro
Preparations for submission to LIPIcs.   - Inline included files.   - Remove commented blocks.   - Update volume macros from portal.

\RequirePackage[hyphens]{url}
\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles.
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

@@

\usepackage{mathtools}
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclareMathOperator{\Rnonneg}{\mathbb R_{\geq 0}}
\DeclareMathOperator{\Ber}{\mathrm{Ber}}
\DeclareMathOperator{\prob}{\mathcal P}
\DeclareMathOperator{\expect}{\mathrm{E}}
\DeclareMathOperator{\indicat}{\mathrm{I}}
\DeclareMathOperator{\bigo}{\mathcal O}
%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

@@

\newcommand\locnew{1003}
\newcommand\locold{2630}

@@

%\title{Verification of the CVM algorithm with a New Recursive Analysis Technique} %TODO Please add
\title{Verification of the CVM algorithm with a Functional Probabilistic Invariant} %TODO maybe?

%\titlerunning{Dummy short title} %TODO optional, please use if title is longer than one line

% TODO: put in everyone's names here
\author{Emin Karayel}{Technical University of Munich, School of Computation, Information and Technology, Germany}{me@eminkarayel.de}{https://orcid.org/0000-0003-3290-5034}{}
\author{Seng Joe Watt}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore}{Watt_Seng_Joe@i2r.a-star.edu.sg}{https://orcid.org/0000-0002-6883-4736}{Singapore NRF Fellowship Programme NRF-NRFF16-2024-0002}
\author{Derek Khu}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore}{derek_khu@i2r.a-star.edu.sg}{https://orcid.org/0009-0000-0293-0664}{}
\author{Kuldeep S. Meel}{Georgia Institute of Technology, United States \and  University of Toronto, Canada}{meel@cs.toronto.edu}{https://orcid.org/0000-0001-9423-5270}{Natural Sciences and Engineering Research Council of Canada (NSERC), funding reference [RGPIN-2024-05956]}
\author{Yong Kiam Tan}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore \and Nanyang Technological University, Singapore}{yongkiam.tan@ntu.edu.sg}{https://orcid.org/0000-0001-7033-2463}{Singapore NRF Fellowship Programme NRF-NRFF16-2024-0002}

\authorrunning{E. Karayel, S. J. Watt, D. Khu, K. S. Meel, Y. K. Tan} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Emin Karayel, Seng Joe Watt, Derek Khu, Kuldeep S. Meel, Yong Kiam Tan} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

@@

\category{} %optional, e.g. invited paper

\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

\supplement{Isabelle/HOL Formalization}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
\supplementdetails[cite=CVM_Distinct_Elements-AFP]{Software}{https://isa-afp.org/entries/CVM_Distinct_Elements.html}
\supplementdetails[cite=CVM_Transforms-Github, swhid=swh:1:dir:c677f1e3ced670373c05ac835f1cd342a2777e20]{Software}{https://github.com/joewatt95/CVM/tree/main/isabelle/CVM_Transforms}
\supplementdetails[cite=Negative_Association-AFP]{Software}{https://isa-afp.org/entries/Negative_Association.html}

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

%\acknowledgements{I want to thank \dots}%optional

%\nolinenumbers %uncomment to disable line numbering

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%NOTE: we have to inline all of these into one file for the final version if accepted.

\input{sec/abstract}

\input{sec/intro}

\input{sec/semra}

\input{sec/invariants}

\input{sec/negdep}

\input{sec/formalization}

\input{sec/formalization_neg_dep}

\input{sec/original_proof}

\input{sec/related_work}

\input{sec/conclusion}

%%
%% Bibliography
%%

%% Please use bibtex,

>>>>

\RequirePackage[hyphens]{url}
\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}

@@

\usepackage{mathtools}
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclareMathOperator{\Rnonneg}{\mathbb R_{\geq 0}}
\DeclareMathOperator{\Ber}{\mathrm{Ber}}
\DeclareMathOperator{\prob}{\mathcal P}
\DeclareMathOperator{\expect}{\mathrm{E}}
\DeclareMathOperator{\indicat}{\mathrm{I}}
\DeclareMathOperator{\bigo}{\mathcal O}

@@

\newcommand\locnew{1003}
\newcommand\locold{2634}

@@

\title{Verification of the CVM algorithm with a Functional Probabilistic Invariant}

\author{Emin Karayel}{Technical University of Munich, School of Computation, Information and Technology, Germany}{me@eminkarayel.de}{https://orcid.org/0000-0003-3290-5034}{}
\author{Seng Joe Watt}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore}{Watt_Seng_Joe@i2r.a-star.edu.sg}{https://orcid.org/0000-0002-6883-4736}{Singapore NRF Fellowship Programme NRF-NRFF16-2024-0002}
\author{Derek Khu}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore}{derek_khu@i2r.a-star.edu.sg}{https://orcid.org/0009-0000-0293-0664}{}
\author{Kuldeep S. Meel}{Georgia Institute of Technology, United States \and  University of Toronto, Canada}{meel@cs.toronto.edu}{https://orcid.org/0000-0001-9423-5270}{Natural Sciences and Engineering Research Council of Canada (NSERC), funding reference [RGPIN-2024-05956]}
\author{Yong Kiam Tan}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore \and Nanyang Technological University, Singapore}{yongkiam.tan@ntu.edu.sg}{https://orcid.org/0000-0001-7033-2463}{Singapore NRF Fellowship Programme NRF-NRFF16-2024-0002}

\authorrunning{E. Karayel, S. J. Watt, D. Khu, K. S. Meel, Y. K. Tan}

\Copyright{Emin Karayel, Seng Joe Watt, Derek Khu, Kuldeep S. Meel, Yong Kiam Tan}

@@

\category{}
\relatedversion{}

\supplement{Isabelle/HOL Formalization}

\supplementdetails[cite=CVM_Distinct_Elements-AFP]{Software}{https://isa-afp.org/entries/CVM_Distinct_Elements.html}
\supplementdetails[cite=CVM_Transforms-Github, swhid=swh:1:dir:c677f1e3ced670373c05ac835f1cd342a2777e20]{Software}{https://github.com/joewatt95/CVM/tree/main/isabelle/CVM_Transforms}
\supplementdetails[cite=Negative_Association-AFP]{Software}{https://isa-afp.org/entries/Negative_Association.html}

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{Yannick Forster and Chantal Keller}
\EventNoEds{2}
\EventLongTitle{16th International Conference on Interactive Theorem Proving (ITP 2025)}
\EventShortTitle{ITP 2025}
\EventAcronym{ITP}
\EventYear{2025}
\EventDate{September 29--October 2, 2025}
\EventLocation{Reykjavik, Iceland}
\EventLogo{}
\SeriesVolume{352}
\ArticleNo{34}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
Estimating the number of distinct elements in a data stream is a classic problem with numerous applications in computer science.
We formalize a recent, remarkably simple, randomized algorithm for this problem due to Chakraborty, Vinodchandran, and Meel (called the CVM algorithm).
Their algorithm deviated considerably from the state of the art, due to its avoidance of intricate derandomization techniques, while still maintaining a close-to-optimal logarithmic space complexity.

Central to our formalization is a new proof technique based on functional probabilistic invariants, which allows us to derive concentration bounds using the Cram\'{e}r--Chernoff method without relying on independence.
This simplifies the formal analysis considerably compared to the original proof by Chakraborty et al.
Moreover, our technique opens up the possible algorithm design space; we demonstrate this by introducing and verifying a new variant of the CVM algorithm that is both total and unbiased---neither of which is a property of the original algorithm.
In this paper, we introduce the proof technique, describe its use in mechanizing both versions of the CVM algorithm in Isabelle/HOL, and present a supporting formalized library on negatively associated random variables used to verify the latter variant.
\end{abstract}

\section{Introduction}
\label{sec:intro}

In 2022, Chakraborty, Vinodchandran, and Meel~\cite{chakraborty2022} published a remarkable streaming algorithm for the distinct elements problem~\cite{quanta}.
Indeed, Knuth later wrote a note on the algorithm~\cite{knuthnote}, pointing out its interesting properties and christening it the \emph{CVM} algorithm (which we use for the rest of this paper).
One striking property of the CVM algorithm is that, in contrast to every other known algorithm for the problem, it does not rely on hashing the stream elements.
Instead, the algorithm could theoretically be implemented in a setting where objects in the data stream only allow for equality comparisons.
Another property is its simplicity, which is why the authors called it ``an algorithm for the textbook''.
The algorithm is displayed in its entirety in~\cref{alg:cvm}.

\begin{algorithm}[h!]
	\caption{CVM algorithm for distinct elements estimation~\cite{chakraborty2022}.}\label{alg:cvm}
	\begin{algorithmic}[1]
  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
  \For{$i \gets 1$ to $l$}
    \State $b \getsr \Ber(p)$ \Comment randomly sample a bit $b$ from the Bernoulli distribution
    \If{$b$} \Comment insert $a_i$ if $b$ is true (with prob. $p$)
      \State $\chi \gets \chi \cup \{a_i\}$
    \Else \Comment remove $a_i$ otherwise
      \State $\chi \gets \chi - \{a_i\}$
    \EndIf
    \If{$|\chi| = n$} \Comment if buffer $\chi$ is full
      \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment discard elements of $\chi$ independently with prob. $\frac{1}{2}$
      \State $p \gets \frac{p}{2}$
    \EndIf
    \If{$|\chi| = n$}
      \Return $\bot$ \Comment fail if $\chi$ remains full
    \EndIf
  \EndFor
  \State \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
  \end{algorithmic}
\end{algorithm}

The pen-and-paper analysis of CVM~\cite{chakraborty2022,chakraborty2023} relies on a sequence of transformations of the algorithm.
The reason for these transformations is that standard methods for analyzing randomized algorithms, such as Chernoff--Hoeffding bounds, usually make statements about independent random variables.
However, for \cref{alg:cvm}, the state variables are far from being independent.\footnote{There is an incorrect claim in the initial published proof of CVM~\cite[Claim 6]{chakraborty2022} that the indicator functions for elements in $\chi$ are independent; a later version by the same authors~\cite{chakraborty2023} provides a correct proof.
The original error serves as a side motivation for this work.}
For example, in Line~3 the Bernoulli distribution is sampled with the parameter $p$, which itself depends on previous random operations; similarly, the subsampling step in Line~9 is only applied if the buffer $\chi$ is full, which also depends on previous random operations.
The aforementioned sequence of transformations by Chakraborty et al.~results in another randomized algorithm which can be analyzed using standard methods, and from which the desired results for the original algorithm can be deduced.
To our knowledge, it seems impossible to analyze \cref{alg:cvm} more directly using known textbook methods~\cite{alon2000,mitzenmacher2017,motwani1995}.

In this paper, we present a new technique for analyzing randomized algorithms which yields a direct and substantially more general proof of the CVM algorithm.
Our approach is very similar to how deterministic algorithms are verified using loop invariants.
The key difference is that our choice of ``loop invariant'' for the randomized streaming algorithm is a functional probabilistic inequality, namely, we consider invariants of the form:
\[
  \expect [ h ] \leq h(c)
\]
where the expectation is taken over the distribution of the state of the algorithm; $h$ is allowed to range over a class of functions (mapping states to real values); and $c$ is a fixed state (possibly chosen differently at each loop iteration).
By first establishing such an invariant for \cref{alg:cvm}, we can then use it (via different choices of $h$) to establish error bounds for the algorithm.
We coined the term ``functional probabilistic invariant'', borrowing loosely from the theory of the calculus of variations, where scalar-valued maps---called functionals---from the problem space are used to solve optimization problems.
In our case we are using scalar-valued maps---i.e., functionals---from the distribution of the state of the algorithm, which lead us to the name for our new technique.
We believe the new proof remains accessible at the undergraduate level, albeit with some exposure to mechanized theorem proving.

To show the generality of our technique, we introduce a new variant of the CVM algorithm, where the subsampling step in Line~9 of \cref{alg:cvm} selects a random $m$-subset of $\chi$ instead of independently discarding each element with some probability.
This variant has the benefit that it is \emph{total} (never returns $\bot$) because the second check in Line~11 becomes obsolete.
More interestingly, the variant is \emph{unbiased}, i.e., the expected value of the algorithm's output is exactly the cardinality of the elements in the stream; this is a new property that neither the original CVM algorithm nor classic algorithms for the distinct elements problem possess.

The modified subsampling step leads to additional dependence for the elements in $\chi$ which cannot be readily removed using transformations as was done in the original proof.
Instead, we verify the new variant with our probabilistic invariant-based approach, using results from the theory of negatively associated random variables~\cite{joagdev1983} to establish the desired functional invariant.
The concept of negative association is a generalization of independence; importantly, negatively associated variables observe closure properties and fulfill Chernoff--Hoeffding bounds similarly to independent random variables.
It should be stressed that the theory of negative association is orthogonal to our new technique, but its formalization is also a contribution of this work.

In summary, our main contributions are:
\begin{itemize}
\item Introduction of a new technique using functional probabilistic invariants to verify tail-bounds for randomized algorithms inductively/recursively.
\item Verification of the original CVM algorithm using our new technique.
\item Presentation and verification of a new variant of CVM that is total and unbiased.
\item Formalization of a theory of negatively associated random variables used to analyze the new CVM variant.
\end{itemize}

We carried out the mechanizations using Isabelle/HOL~\cite{nipkow2002}, which comes with a large repository of foundational libraries~\cite{afp} for the verification of randomized algorithms.
We have also mechanized the transformation-based CVM proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}, which provides a rough point of comparison:
verification of the CVM algorithm using our new technique required only \locnew~lines, while the original proof required \locold~lines.\footnote{
We count the total number of lines of Isabelle code in the whole project, excluding empty, comment, and presentation-related lines.
}

The rest of this paper is organized as follows.
\cref{sec:background} provides background information on randomized algorithms, in particular on their semantics in Isabelle/HOL.
\cref{sec:invariants} introduces our new technique and explains how probabilistic loop invariants can be used to establish tail bounds for the original CVM algorithm.
\cref{sec:negdep} introduces the concept of negative association and our new total and unbiased variant of the CVM algorithm.
\cref{sec:formalization} presents the formalization of both variants of the algorithm, and \cref{sec:formalization_neg_dep} describes our new formalized library on negatively associated random variables.
\cref{sec:transformation_based_proof} discusses some challenges faced in our alternative verification of CVM using the transformation-based proof by Chakraborty et al.
The final sections present related work and a summary of our results.

The supplementary material contains:
\begin{itemize}
\item formalization~\cite{CVM_Distinct_Elements-AFP} of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) using functional probabilistic invariants;
\item formalization~\cite{Negative_Association-AFP} of a library for negative association; and
\item formalization~\cite{CVM_Transforms-Github} of the CVM algorithm following the proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}.
\end{itemize}

\section{Background}
\label{sec:background}

\subsection{Randomized Algorithms and Distinct Elements}

The CVM algorithm is a \emph{streaming} algorithm for the distinct elements problem.
As shown in~\cref{alg:cvm}, given a data stream $a_1,\dots, a_l$, the goal of such algorithms is to return an accurate cardinality estimate for the set $A = \{a_1,\dots,a_l\}$.

Importantly, CVM is a \emph{probably approximately correct} (PAC) algorithm where its output estimate $R$ satisfies
$\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
for parameters $\varepsilon$ and $\delta$,
i.e., the probability that the relative error of $R$ with respect to $|A|$ exceeds $\varepsilon$ is at most $\delta$.
Moreover, let us assume that the space needed to store each element in the stream is $b$ bits, then the CVM algorithm requires only $\bigo(\varepsilon^{-2} b \ln(\delta^{-1} l))$ bits of mutable state, which is far less than storing each stream element deterministically.

\begin{remark}
The asymptotically optimal randomized algorithm for distinct elements requires $\bigo( \varepsilon^{-2} \ln \delta + b)$ bits, but it requires more advanced algorithmic techniques. It would not be possible to present using such elementary steps as in \cref{alg:cvm} as it involves computations in finite fields and random walks in expander graphs~\cite{blasiok2020, karayel2023}.
\lipicsEnd\end{remark}

\subsection{Semantics of Randomized Algorithms}
We briefly review how reasoning about randomized algorithms works in Isabelle/HOL using the Giry monad~\cite{giry1982}.
Multiple authors provide more thorough discussions of the concept in the context of Isabelle and other proof assistants~\cite{audebaud2009,eberl2020, lochbihler2016}.

The key idea is to model a randomized algorithm as a probability space representing the distribution of its results.
As an example, let us consider \cref{alg:example}.
\begin{algorithm}[h!]
\caption{Example for sequential composition.}\label{alg:example}
\begin{algorithmic}[1]
\State $p \getsr \Ber(\frac{1}{2})$
\State $q \getsr \Ber(\frac{1}{3}+\frac{p}{2})$
\State \Return $q$
\end{algorithmic}
\end{algorithm}%

In the first step,~\cref{alg:example} flips a fair coin, such that $p$ is $1$ with probability $\frac{1}{2}$ and $0$ otherwise; the notation $\Ber(p)$ represents the Bernoulli distribution.
In the second step, the algorithm flips a coin $q$ which depends on $p$.
This has the consequence that, to semantically model $q$, we have to consider functions returning probability spaces, like: $p \mapsto \Ber(\frac{1}{3}+\frac{p}{2})$, which is being \emph{bound} to the distribution of $p$.
The resulting distribution for $q$ is a \emph{compound distribution} resulting from a combination of $\Ber(\frac{1}{3})$ (when $p = 0$) and $\Ber(\frac{5}{6})$ (when $p = 1$).

This example captures the main aspects of modeling randomized algorithms in the Giry monad.
Indeed, randomized algorithms can be modeled using the following ingredients:

\begin{description}
\item[Primitive Random Operations.] For example, a simple fair coin flip is represented using the Bernoulli distribution, $\Ber(\frac{1}{2})$.
\item[Return Combinator.]
Given an element $x$, we can construct the singleton probability space, assigning probability $1$ to $x$ and $0$ to everything else.
In monad notation, this is written as: $\mathrm{\bf return}\, x$.

\item[Bind Combinator.]
The bind combinator represents sequential composition of two randomized algorithms $m$ and $f$, where the latter randomized algorithm consumes the output of the former; in monad notation, this is: $m \isa{\isasymbind} f$.
Mathematically, this is the most involved operation, because $f$ is a function returning probability spaces, which takes inputs from the probability space $m$.

Let us consider an event $A$ in the probability space $m \isa{\isasymbind} f$.
Its probability can be evaluated by integrating over its probabilities in $f$ with respect to $m$:
\[
  \prob_{m \isa{\isasymbind} f} (A) = \int_m \prob_{f(x)} (A) \, d x \textrm{.}
\]

Another key property is the calculation of expectations;
if $h$ is a random variable over $m \isa{\isasymbind} f$, we can compute its expectation as:
\begin{equation}
  \label{eq:integral_bind}
  \expect_{m \isa{\isasymbind} f} [h] = \int_m \expect_{f(x)} [h] \, d x \textrm{.}
\end{equation}

Equation~\ref{eq:integral_bind} is crucially used to establish the invariants we introduce next in~\cref{sec:invariants}.
\end{description}

\section{Functional Probabilistic Invariants}\label{sec:invariants}
In this section, we will derive our new technique using \cref{alg:cvm} as an example.
Let us start by briefly reviewing the algorithm---its state is a buffer $\chi$ (initially empty) and a fraction $p > 0$ (initially set to $1$).
The buffer tracks a subset of the elements of the stream encountered so far, with maximal size $n$ chosen according to the desired accuracy parameters $\varepsilon$, $\delta$, and the stream size $l$.
The algorithm iterates over the stream elements, adding each one to the buffer with probability $p$ or conversely---if the current stream element is already in the buffer---removing it with probability $(1-p)$ (Lines 3--7).
If the number of elements in the buffer reaches the maximal size $n$, the subsampling operation is executed, which discards each element in $\chi$ independently with probability $\frac{1}{2}$; then, $p$ is adjusted to reflect the fact that the buffer now contains each element with probability $p_\text{new} = \frac{p_\text{old}}{2}$ (Lines 8--10).
If the subsampling operation fails, i.e., if no elements get discarded, then the algorithm fails returning $\bot$ (Line 11).
After processing the stream, the algorithm returns $\frac{|\chi|}{p}$ as a probably-approximately correct estimate for the number of distinct elements in the stream.

\begin{remark}
For our discussion below, it is convenient to analyze \cref{alg:cvm} without Line 11, i.e., we will skip the second check of $|\chi|=n$ determining whether the subsampling step succeeded.
This modified version simplifies our analysis as we do not have to worry about the possibility of the algorithm failing (returning $\bot$).
This transformation is also used in the original CVM proof~\cite{chakraborty2023}, where the total variational distance between these two variants of the algorithms is shown to be at most $\frac{\delta}{2}$.
Thus, probability bounds derived for the modified version can be transferred to the original algorithm, with a correction term of $\frac{\delta}{2}$.
\lipicsEnd\end{remark}

\subsection{Deriving a Simple Probabilistic Invariant}
Consider the random variables $X_s := \indicat(s \in \chi)$ indicating the presence of a stream element $s \in A = \{a_1,\ldots,a_l\}$ in the buffer, where we write $\indicat$ for the indicator of a predicate, so $\indicat(\mathrm{true}) = 1$ and $\indicat(\mathrm{false}) = 0$.
Before the algorithm first encounters the stream element $s$, $X_s$ will be $0$ unconditionally, because the buffer $\chi$ is always a subset of the stream elements processed so far, i.e., $\chi \subseteq \{a_1,\dots,a_m\}$ after loop iteration $m$.

In the loop iteration where element $s$ occurs for the first time, it will be inserted with probability $p$ in Lines 3--7.
This means, after Line 7, we have:
\begin{equation}
  \label{eq:indicator_eq}
  \expect [p^{-1} X_s] = 1 \textrm{.}
\end{equation}
Interestingly, this equation is preserved for the rest of the algorithm.
For example, let us consider a subsampling step: each $s$ is independently discarded with probability $\frac{1}{2}$ so $\prob(X_s=1)$ is halved, but so is $p$ after subsampling, which preserves the equation.

Let us see how we can verify Equation~\ref{eq:indicator_eq} more formally.
For that, we model the state of the randomized algorithm as a pair $(\chi,p)$ and we write $\chi$ and $p$ for the random variables projecting their respective components from the distribution of the state of the algorithm.
We will refer to parts of each loop iteration in \cref{alg:cvm} as $\mathrm{step}_1$ (resp.~$\mathrm{step}_2$) for Lines 3--7 (resp.~Lines 8--10).
The final distribution of the algorithm is the distribution resulting from the sequential composition of alternating steps over the stream:
\[
  \mathrm{init} \, \isa{\isasymbind}\; \mathrm{step}_1\, a_1 \; \isa{\isasymbind}\; \mathrm{step}_2 \; \isa{\isasymbind}\; \mathrm{step}_1 \, a_2 \; \isa{\isasymbind}\; \cdots \isa{\isasymbind}\; \mathrm{step}_1\, a_l \; \isa{\isasymbind}\; \mathrm{step}_2
\]
where we parameterize $\mathrm{step}_1$ with the stream element that it processes.
The term $\mathrm{init}$ represents the initial state, i.e., $\mathrm{init} = \mathrm{\bf return}\,(\{\},1)$.
It is easy to show by induction over the sequence of steps, we have $0 < p \leq 1$ and $\chi \subseteq A$ for all possible states of the algorithm.

Let us verify that Equation~\ref{eq:indicator_eq} is preserved as an invariant over all steps.
To verify that $\mathrm{step}_1\, a$ preserves Equation~\ref{eq:indicator_eq}, we assume some probability space of states $\Omega$ fulfills Equation~\ref{eq:indicator_eq} and we would like to show that it is still true for $\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a$. By Equation~\ref{eq:integral_bind},
\begin{equation}\label{eq:step_1_exp}
  \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a} [ p^{-1} X_s ] =
    \integral{\Omega}{\integral{\Ber(p)}{p^{-1} \indicat\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right)}{\tau}}{\sigma} \textrm{.}
\end{equation}
Note that we write $p$ or $\chi$ even though we should actually write $p(\sigma)$ or $\chi(\sigma)$, i.e., we remember that these implicitly depend on $\sigma$.
To see that the right-hand side is equal to $1$, it is useful to consider cases on whether $a=s$.
When $a = s$, the right-hand-side is equal to $1$ by definition of the Bernoulli distribution (since $p \in (0;1]$).
When $a \not= s$, it follows from the induction hypothesis on $\Omega$; in particular, the term in the inner integral is constant with respect to $\tau$.

The same invariant-based argument is possible for $\mathrm{step}_2$.
Let us assume $\Omega$ is a probability space of states fulfilling Equation~\ref{eq:indicator_eq}.
Then by Equation~\ref{eq:integral_bind}, $\expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} [p^{-1}X_s]$ equals
\begin{equation}\label{eq:step_2_exp}
    \integral{\Omega}
      {\left(\ift{|\chi|=n}{\left(\integral{\mathrm{subsample}(\chi)}{\frac{\indicat(s \in \tau)}{p/2}}{\tau}\right)}{\frac{\indicat(s \in \chi)}{p}} \right)}
      {\sigma}
    \textrm{.}
\end{equation}
Note that the \emph{true} and \emph{false} cases of the inner \textbf{if-then-else} both evaluate to the same value: $p^{-1} \indicat(s \in \chi)$.
If $s \notin \chi$ both sides of the equation are $0$, because the subsampling operation returns a subset of $\chi$.
If $s \in \chi$ the probability that the element gets subsampled is $1/2$, so we arrive again at $\frac{1/2}{p/2} = p^{-1} \indicat(s \in \chi)$.
Hence: $\expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} [p^{-1} X_s] = \expect_\Omega [p^{-1} X_s] = 1$.
This completes the invariance proof for Equation~\ref{eq:indicator_eq}.

\subsection{Deriving a Functional Probabilistic Invariant}
With Equation~\ref{eq:indicator_eq} established, it is straightforward to show that the expected value of the output estimate $p^{-1} |\chi|$ for the modified algorithm (without Line 11) is equal to the desired cardinality $|A|$.
However, recall that we are interested in verifying the estimate's PAC guarantee.
A typical approach to establishing such a guarantee is to use Chernoff bounds which provide exponential tail bounds (i.e., concentration bounds) for the deviation of sums of independent random variables from their mean.
However, these are not directly useful in the CVM algorithm because the key random variables, e.g., $p^{-1} X_s$ for $s \in A$, are dependent.

An alternative is the Cram\'{e}r--Chernoff method, which is a general method to obtain tail bounds for any random variable.
It can be stated simply as $\prob(X \geq a) \leq M(t) e^{-ta}$ for all $t > 0$, where $M(t) := \expect [\exp(t X)]$ is the moment generating function of the random variable $X$.
It is also possible to obtain lower tail bounds $\prob(X \leq a)$ using the Cram\'{e}r--Chernoff method, which just requires estimates for $M(t)$ for $t < 0$, instead of $t > 0$.

In our case, we are interested in estimating the moment generating function of the random variable $p^{-1} |\chi|$ for the CVM algorithm:
\[
  \expect [\exp( t p^{-1} |\chi| )] = \expect \left[ \prod_{s \in A} h(p^{-1} X_s) \right]
\]
for $h(x) = \exp(tx)$.
At this point, it is tempting to see whether the proof for Equation~\ref{eq:indicator_eq} can be generalized to establish bounds for the above.
Indeed, we managed to establish the following generalized result:
\begin{align}
  \expect \left[ \prod_{s \in A} h(p^{-1} X_s) \right] \leq h(1)^{|A|} \label[ineq]{i:func_invariant}
\end{align}
for every non-negative concave function $h : \Rnonneg \rightarrow \Rnonneg$.
However, the exponential function in $M(t)$ is convex, but we can instead try to derive tail bounds for the random variable $\indicat(p \geq q) p^{-1} |\chi|$, for some fixed constant $q > 0$.
This leads to a similar invariant inequality:
\begin{align}
  \expect \left[ \prod_{s \in A} I(p \geq q) h(p^{-1} X_s) \right] \leq h(1)^{|A|} \label[ineq]{i:func_invariant_capped}
\end{align}
with the new condition that $h$ needs to be non-negative and concave only on $[0;q^{-1}]$.
This then allows us to approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which yields tail bounds for $p^{-1} |\chi|$ under the condition $p \geq q$. As an example, the upper tail bound can be derived as follows:
\begin{eqnarray*}
  \prob( p^{-1} |\chi| \geq (1+\varepsilon) |A| \wedge p \geq q) & \leq & % Is not equal when |A|=0.
  \prob( \indicat(p \geq q) p^{-1} |\chi| \geq (1+\varepsilon) |A|) \\
  & \underset{\textrm{Markov}}{\leq} & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} \indicat(p \geq q) \exp( t p^{-1} X_s) \right] \\
  & \leq & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} \indicat(p \geq q) h(p^{-1} X_s) \right] \\
  & \underset{\textrm{Ineq.~\ref{i:func_invariant_capped}}}{\leq} & e^{-t(1+\varepsilon) |A|} h(1)^{|A|} \\
  & \underset{\textrm{Calculus}}{\leq} & e^{-n\varepsilon^2 / 12}
\end{eqnarray*}
where we choose $h(x) = 1+qx (e^{t/q}-1)$.
Note that $h$ is affine and it can be easily checked%
\footnote{Because the exponential function is convex and $h$ is affine, we only have to check the end points: $0, q^{-1}$.}
that it is an upper approximation of $\exp(tx)$ for $x \in [0;q^{-1}]$.
For the last step, we have to find the $t$ that produces the required bound.%
\footnote{We use $t = q \ln(1+\varepsilon)$ which is not the real optimum, but better for algebraic evaluation.}
To use these bounds, we also have to separately estimate $\prob(p < q)$.
For that, we use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2023}, with $q = \frac{n}{4 |A|}$.
The formalization in the supplementary material~\cite{CVM_Distinct_Elements-AFP} contains a detailed informal step-by-step proof using our approach in its appendix.
Besides the use of Equation~\ref{eq:integral_bind} and the Cram\'er--Chernoff method, the steps are elementary.

We call inequalities like Inequality~\ref{i:func_invariant} and \ref{i:func_invariant_capped}: \emph{functional probabilistic invariants}.
The inequalities can be established using induction over the steps of the randomized algorithm.
Even, though, the actual distribution of the states themselves are not known, nor do we think it is possible to find useful closed form descriptions for them, the invariant establishes valuable information about the distribution of the state. In this case, enough information, to establish exponential tail bounds for the algorithm.

\section{An Unbiased CVM Variant and Negative Dependence\label{sec:negdep}}
An interesting consequence of our invariant-based approach is that it allowed us to devise and verify a refined version of the CVM algorithm that is both total and unbiased.
